### 2022-06-23 manuscript/response_to_reviewers.tex

    \textcolor{blue}{We re-ran the simulations to assess the false positive rate and include the results in the Materials and Methods section: ``For our main results (Figure 2) we simulated 20 truly associated variants, as described above. To also check the false positive rate (FPR), we re-ran the simulations with an additional 80 non-associated variants. Across all the association tests in this second simulation setup (7 $H^2$ levels × 10 $\alpha$ levels × 100 variants × 20 replicates per scenario = 140,000 association tests), FPR was 0.0005 using the true (simulated) non-pathogen part of the trait, 0.0005 using the estimated non-pathogen part of the trait, and 0.0006 using the scaled total trait value. These rates are comparable to the expected FPR of 0.0005 at significance level 0.05 corrected for 100 tests. Given the stricter correction for multiple testing in this second simulation setup, the TPR decreased significantly across all three GWAS response variables used.''}
    \item I don't think panel A of figure 2 is explained well by the caption or text, I could not understand what the RMSE is showing here. What does 'more accurate' look like in this plot? \\
    \textcolor{blue}{We edited the figure legend to make it clearer: ``(A) shows the root mean squared error (RMSE) of our estimator (left) compared to un-corrected trait values, scaled by their mean (right) under each simulated evolutionary scenario. The RMSE is with reference to the true (simulated) host part of the trait values. Thus, more accurate estimates (lower RMSE) mean the trait value used for GWAS will be closer to the true host part of the trait value.''}
    \item Is there any explanation why the simulated (actual) and estimated non-pathogen trait in fig 2B start to look quite different at stronger selection strengths. \\
    \textcolor{blue}{We added a figure (Figure 4) to illustrate how the correlation between clustered tips on the phylogenetic tree decreases with increasing selection strength/ increasing stochasticity. Without phylogenetic signal, our method has no information to estimate the pathogen effect on the trait. This is described in the section ``Estimator accuracy'': ``Depending on these parameters, trait values are more or less phylogenetically correlated (see also Figure 4) and the phylogeny is more or less useful for accurately estimating the heritable pathogen and corresponding non-heritable, non-pathogen part of the trait values.''}
    \item Using plink and removing samples with divergent ancestry is probably a little outdated for host GWAS. A linear mixed model would likely provide more power still. In particular, I'd suggest warped-lmm (http://dx.doi.org/10.1038/ncomms5890), which also transforms the phenotype to increase power. Furthermore, I would be interested to see histograms/scatter plots of a few phenotypes: a) no adjustment; b) warped-lmm only; c) the author's method; d) both adjustments. Perhaps the methods are similar, or using both does increases power beyond either method alone. \\
    \textcolor{blue}{We highlight that our method is a pre-processing step before GWAS and can be combined with many different GWAS models implemented in softwares like PLINK, GEMMA, and warped-lmm. Our revised manuscript focuses more on the method rather than generating biological insights for HIV or \emph{X. arboricola}, so we feel that updating our example GWAS is beyond the scope of the current study. We added this explanation to the discussion: ``Our approach is envisioned to be complementary to previously developed methods for infectious disease GWAS.... In cases where a correction can be estimated and applied, the corrected trait values are envisoned to be used in any of the previously developed GWAS models for the actual association testing (we used a linear model approach implemented in PLINK (Chang et al., 2015), though a more advanced method would be to use a linear mixed model with host ancestry as a random effect).''}
\end{enumerate}

\vspace{5mm}
Suggested text changes:

\begin{enumerate}
    \item The first sentence of the abstract is probably also true for many common diseases analysed in human GWAS (apart from perhaps the pathogen, but gut microbiome may be a factor too). \\
    \textcolor{blue}{We changed the wording to make it clearer what sets infectious disease GWAS apart: ``Infectious diseases are particularly challenging for genome-wide association studies (GWAS) because genetic effects from two organisms (pathogen and host) can influence a trait.''}
    \item L84-86. There are more than two methods for this, see also https://doi.org/10.1371/journal.pgen.1007758; https://doi.org/10.1093/bioinformatics/bty539; https://doi.org/10.1099/mgen.0.000469. Maybe remove 'two', as this will change anyway.\\
    \textcolor{blue}{We removed the ``two'' as suggested: ``More recently, analogous methods have been developed for microbial GWAS, where clonal reproduction exacerbates population structure (Power et al., 2017). Microbial GWAS-specific phylogenetic methods to account for population structure in microbial GWAS include explicitly testing for lineage-specific effects as in (Earle et al., 2016) and modified association tests that account for phylogenetic relationships amongst samples as in (Collins and Didelot, 2018).''}
    \item L110 'Of these, only pathogen effects are heritable from one transmission partner to another'. This might be nit-picking but I could see this phrasing being misunderstood by some readers. Environment will be partially shared along a transmission chain, probably. Interaction effects could also be shared if e.g. hosts are likely to have the same ancestry.\\
    \textcolor{blue}{We improved the phrasing to be more accurate: ``GWAS typically stratify samples or include covariates to correct for host genetic factors or non-genetic factors that may be correlated with a trait value. This leaves pathogen genetic factors as a remaining source of correlation, since close transmission partners may be infected with very similar pathogen strains.''}
    \item Equation 1 onwards. Using 'e' for the environmental effects is easy to get confused with more standard use of the number e = 2.718, especially when raised to a power. Consider changing this symbol.\\
    \textcolor{blue}{We changed the variable to $\epsilon$ and also changed the variable for the pathogen part of the trait to $g$ to match previous work on the POUMM (e.g. \citep{Mitov2017a-POUMM})}.
    \item Please add the link to the github code in the framework section at the end of the methods.\\
    \textcolor{blue}{We added the link in the section ``Phylogenetic trait correction'': ``All the code used to implement the method is available at https://github.com/cevo-public/POUMM-GWAS.''}
    \item L283 'In simulations, we showed this is the case when heritability is high, selection strength is low, and trait values are not subject to strong stochastic fluctuations.' Could the authors give a biological example of this? Would antimicrobial resistance variation fit the bill?\\
    \textcolor{blue}{We have not fit the POUMM to data from enough systems to have a clear example of this, so instead we suggest other systems that might fall into this case: ``Future applications of our method might investigate other clinically significant disease traits and outcomes that are affected by both host and pathogen genetic factors, for instance Hepatitis B Virus-related hepatocellular carcinoma (An et al., 2018), Hepatitis C treatment success (Ansari et al., 2017), and susceptibility to or severity of certain bacterial infections, e.g. Donnenberg et al. (2015); Messina et al. (2016). Transcriptomic data has also previously been modeled as an evolving phenotype using an Ornstein-Uhlenbeck model (Rohlfs et al., 2014). Thus, one could also estimate pathogen effects on host gene expression.''}

    \textcolor{blue}{We anticipate antimicrobial resistance would be under strong selective pressure, but agree this could produce phylogenetically correlated trait values. We added a statement in the discussion: ``It is possible that traits under strong selection pressure are also highly heritable and phylogenetically correlated, for instance antimicrobial resistance. However, such traits might violate the POUMM assumption that trait values vary as a random walk in continuous space if they are caused by few mutations of strong affect.''}
\end{enumerate}
\end{document}
